---
title: "Practical Machine Learning assignment"
author: "maisam"
date: "5/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Over View
This document is the final report of the Peer Assessment project from Coursera’s course Practical Machine Learning, as part of the Specialization in Data Science. It was built up in RStudio, using its knitr functions, meant to be published in html format.
This analysis meant to be the basis for the course quiz and a prediction assignment writeup. The main goal of the project is to predict the manner in which 6 participants performed some exercise as described below. This is the “classe” variable in the training set. The machine learning algorithm described here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.

## Background
This project aims to predict in which manner participants completed a single exercise, based on recordings provided by accelerometers attached to 4 measurement points on the body. Six male participants aged 20-28 were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in one correct, and 4 incorrect fashions. Specifically, the exercises were performed:  
* Exactly according to the specification (__Class A__);  
* Throwing elbows to the front (__Class B__);  
* Lifting the dumbbell only halfway (__Class C__);  
* Lowering the dumbbell only halfway (__Class D__); and  
* Throwing the hips to the front (__Class E__).

## Environment Preparation
We first upload the R libraries that are necessary for the complete analysis.

```{r , message = FALSE, cache = TRUE}
rm(list=ls())                # free up memory for the download of the data sets
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)
```
## Data Loading and Cleaning
The next step is loading the dataset from the URL provided above. The training dataset is then partinioned in 2 to create a Training set (70% of the data) for the modeling process and a Test set (with the remaining 30%) for the validations. The testing dataset is not changed and will only be used for the quiz results generation.



```{r , message = FALSE, cache = TRUE}
# download the datasets
training <- read.csv("pml-training.csv")
testing  <- read.csv("pml-testing.csv")

# create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)

```


```{r , message = FALSE, cache = TRUE}
dim(TestSet)
```
Both created datasets have 160 variables. Those variables have plenty of NA, that can be removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.

```{r , message = FALSE, cache = TRUE}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TrainSet)
```


```{r , message = FALSE, cache = TRUE}
# remove variables that are mostly NA
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
dim(TrainSet)
```


```{r , message = FALSE, cache = TRUE}
# remove identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
```
With the cleaning process above, the number of variables for the analysis has been reduced to 54 only.

## Corelation Analysis
A correlation among variables is analysed before proceeding to the modeling procedures.
```{r , message = FALSE, cache = TRUE}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```
The highly correlated variables are shown in dark colors in the graph above. To make an evem more compact analysis, a PCA (Principal Components Analysis) could be performed as pre-processing step to the datasets. Nevertheless, as the correlations are quite few, this step will not be applied for this assignment.

## Predictive Model Building
Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. The methods are: Random Forests and Decision Tree  as described below.
A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.

# a) Random Forest

```{r , message = FALSE, cache = TRUE}
# model fit
set.seed(12345)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
                          trControl=controlRF)
modFitRandForest$finalModel
```


```{r , message = FALSE, cache = TRUE}
# prediction on Test dataset
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest

```


```{r , message = FALSE, cache = TRUE}
# plot matrix results
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 4)))

```
# Decision Tree


```{r , message = FALSE, cache = TRUE}
# model fit
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)
```


```{r , message = FALSE, cache = TRUE}
# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree

```



```{r , message = FALSE, cache = TRUE}
# plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))

```


# Applying the Selected Model to the Test Data

The accuracy of the 2 regression modeling methods above are:

    Random Forest : 0.9963
    Decision Tree : 0.7368

In that case, the Random Forest model will be applied to predict the 20 quiz results (testing dataset) as shown below.
```{r , message = FALSE, cache = TRUE}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```
